Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-02 causal-bandit/icml-2015/general.tex
Replace \leq with = in variance bound.
Updated contributions

Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
We would like to estimate
\eq{
\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
}
using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$. Ultimately we will be interested
in the case that 
$Q(x) = P(x|do(X_i = j))$
and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
\eqn{
\label{eq:truncated}
Z = Y R \ind{R < B}\,,
}
which trivially satisfies $|Z| \leq B$ surely.
Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
Computing the expectation we see
\eq{
\EE_P[Z] = \mu - \beta\,.
}
where $\beta \geq 0$ is the negative bias and satisfies
\eq{
\beta \leq \Q{R \geq B}\,.
}
Bounding the variance is also straightforward.
\eq{
\Var[Z] 
&\leq \EE[Z^2] 
=\sum_x \P{X = x} Z(x)^2 \\
&\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 \\
&\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
= V\,.
}
The bias can be controlled in terms of $V$ via Markov's inequality.
\eq{
\beta 
\leq \Q{R \geq B} 
\leq \frac{\EE_Q[R]}{B}
= \frac{V}{B}\,.
}
The point is that $B$ can be tuned to trade the bias against the concentration of the estimator,
the latter of which will be controlled via Bernstein's inequality.

@@

\todot{discuss role of $delta = 1/(NT)$}
\todot{define $Z_{ijkl}(t)$ etc.}
\todot{Define $B_{ijkl}$}
\begin{theorem}
Let $\epsilon > 0$ be the smallest value of $\epsilon$ such that Algorithm \ref{alg:alloc} returns true and let $IJ = \argmax_{ij} \hat \mu_{ij}$.
Then $\mu^* - \E \mu_{IJ} \leq \epsilon + 1/T$.
\end{theorem}

>>>>

Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
We would like to estimate
\eq{
\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
}
using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$. Ultimately we will be interested
in the case that 
$Q(x) = P(x|do(X_i = j))$
and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
\eqn{
\label{eq:truncated}
Z = Y R \ind{R < B}\,,
}
which trivially satisfies $|Z| \leq B$ surely.
Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
Computing the expectation we see
\eq{
\EE_P[Z] = \mu - \beta\,.
}
where $\beta \geq 0$ is the negative bias and satisfies
\eq{
\beta \leq \Q{R \geq B}\,.
}
Bounding the variance is also straightforward.
\eq{
\Var[Z] 
&\leq \EE[Z^2] 
=\sum_x \P{X = x} Z(x)^2 \\
&\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 \\
&= \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
= V\,.
}
The bias can be controlled in terms of $V$ via Markov's inequality.
\eq{
\beta 
\leq \Q{R \geq B} 
\leq \frac{\EE_Q[R]}{B}
= \frac{V}{B}\,.
}
The point is that $B$ can be tuned to trade the bias against the concentration of the estimator,
the latter of which will be controlled via Bernstein's inequality.

@@

\todot{discuss role of $delta = 1/(NT)$}
\todot{define $Z_{ijkl}(t)$ etc.}
\todot{Define $B_{ijkl}$}
\begin{theorem}\label{thm:general-upper}
Let $\epsilon > 0$ be the smallest value of $\epsilon$ such that Algorithm \ref{alg:alloc} returns true and let $IJ = \argmax_{ij} \hat \mu_{ij}$.
Then $\mu^* - \E \mu_{IJ} \leq \epsilon + 1/T$.
\end{theorem}

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
Updated contributions

Our main results for this case are: an upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?); an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.

Although the witness (Algorithm~\ref{alg:}) for our upper bound is very simple (\ie, observe without intervening for some initial period and then play undersampled actions in order to get good estimates) our key point is that algorithms which use causal information when available have a clear, theoretical improvement over those which do not.
These results are experimentally validated in \S\ref{sec}.

>>>>

In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently.
We propose a simple causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show it is optimal up to log factors (Theorems~\ref{thm:uq-simple} \& \ref{thm:lower}) with $\tilde{\Theta}(\sqrt{m/T})$ simple regret after $T$ rounds where $m$ depends on the casual model and may be much smaller than $N$.
In contrast, the best a bandit algorithm can do if the causal model is ignored is $\Omega(\sqrt{N/T})$.
This shows theoretically the value of our framework over the traditional bandit problem. 
Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in bandit problems.

Interventions and observations may have a complex relationship in the general causal model. 
However, in \S\ref{sec:simple-regret-general}, Theorem~\ref{thm:general-upper} shows how an importance-sampling based algorithm (Algorithm~\ref{alg:alloc}) can obtain non-trivial guarantees on simple regret.

% Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
% Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.

% Although the witness (Algorithm~\ref{alg:}) for our upper bound is very simple (\ie, observe without intervening for some initial period and then play undersampled actions in order to get good estimates) our key point is that algorithms which use causal information when available have a clear, theoretical improvement over those which do not.
% These results are experimentally validated in \S\ref{sec}.
